---
title: "Hotel Bookings"
author: "Nicolas Pacheco Soliz"
date: "8/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Purpose: The purpose of this project is to conduct an EDA of hotel bookings and answer questions given data. Next, we want to attempt to predict whether or not a hotel will be booked using classical machine learning algorithms. 
```{r}

#install.packages("readr")
library(readr)
#install.packages("dplyr")
#install.packages("tidyverse")
library(dplyr)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("lubridate")
library(lubridate)
library(tidyverse)
library(car)


df <- read_csv("hotel_bookings.csv")


##what countries have the most hotel bookings? 
## what are the trends for different hotel bookings? what times of year seem to be most busy?
##high number of cancellations per time of year? 
#bookings by market segment? 
#which month has the highest number of canllations?
##how much do guests pay for a room per night?

##which has more cancellations, resort or ohther hotel? 

####lead times vs canclled density

##we canu sne density curves here!



```



Data Cleaning 
```{r}
str(df)

df$is_canceled <- as.factor(df$is_canceled)
levels(df$is_canceled) <- c("Not Canceled", "Canceled")

##checking for missing values 
df[which(is.na(df)),]  ## only 4 NAs, so we will omit the missing valules 
df<- na.omit(df)##Data without NAs
str(df)


##we also see that companies and agents also have "NULL" values;

nrow(df[which(df$company == "NULL"),])/ nrow(df)  ##94% of these values in this column have missing values, so we will have to delete this column 

df <- df %>% select(-company)

nrow(df[which(df$agent == "NULL"),])/ nrow(df) ##13% of agent rows are missing, so we will investigate this later; 

```


Time Series Analysis
```{r}


df$date <- as.Date(strptime(paste(df$arrival_date_year, df$arrival_date_month,df$arrival_date_day_of_month,sep="-"), format = "%Y-%b-%d"))
timedata <- df %>% group_by(date) %>% count(date)

ggplot(timedata, aes(x=date, y=n)) + geom_line() +  xlab("") + ggtitle("Number of Bookings Over Time") + theme(plot.title = element_text(hjust = 0.5))



ggplot(data = df, aes(x = arrival_date_month)) + geom_bar(aes(fill = is_canceled), width = 0.5) + ggtitle("Bookings and Cancelletations Per Month") + theme(plot.title = element_text(hjust = 0.5)) + coord_flip()


```

Cancelations By Month

```{r}

#order <- df %>% group_by(arrival_date_month) %>% count(arrival_date_month) %>% arrange(desc(n)) ##if we want to make it by order, we can these two lines of code 
#df$arrival_date_month  <- factor(df$arrival_date_month, levels = order$arrival_date_month)

###here we see bookings and cancellations per monts; 

df$arrival_date_month  <- factor(df$arrival_date_month, levels =c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) ##Reordering the months 
ggplot(data = df, aes(x = arrival_date_month)) + geom_bar(aes(fill = is_canceled), width = 0.5) + ggtitle("Bookings and Cancelletations Per Month") + theme(plot.title = element_text(hjust = 0.5)) + coord_flip() 

##from the plot, we can see that the higher the number of bookings, the higher the cancellation rate; they tend to move with each other; if this is the case, and the proportion of cancelations is relatively constant, then taking into account month might provide strong predictive power for our model. 


#let's find the cancellation rates per month, and figure out all months are proportional or some have a higher cancelations than others 
perc_cancellations <- df %>% group_by(arrival_date_month) %>% summarise(perc = mean(is_canceled))
mean(df$is_canceled)  ##Overall cancellation rate is 37%


ggplot(data = perc_cancellations) +  geom_bar(aes(x = arrival_date_month, y = perc, fill = perc), stat = 'identity') +  scale_fill_gradient(high = "#132B43",
  low = "#56B1F7") + ggtitle("Percent Cancelations By Month") + theme(plot.title = element_text(hjust = 0.5)) ## people are most likely to cancel in the beginning of the year 
##from the plot above we can see that the higher the cancelation months tend to be april and june; these dates mark spring break and beginning of summer, so this is likely why we tend to see a higher conentration of canelleations here 






```


```{r}

ggplot(data =df, aes(x = lead_time)) + geom_density(aes(fill = is_canceled),  stat = 'density', alpha= 0.5) + ggtitle("\tFrequency of Cancellations Given Lead Times") + theme(plot.title = element_text(hjust = 0.5))

p = ggplot(data = df)


##We can see that most of the cancellations happened with low lead time; there is an obvious relationship between lead time and nubmer of cancellations; 

```


From this point, we can definitely include the month and cancelation in our model; let's identify some other relationships 


What countries book tend to account for most of the bookings? Which countries have the highest number of cancelations? 

```{r}



###how likely are repeated guests to cancel? 
tab <- table(hotels$is_canceled, hotels$is_repeated_guest)
tab
##given that a guest cancels, it is is more likely they are NOT a repeated guest; only 552 guests that canceled were repeated guests; roughly 96.5% of guests who cacneled where non returning, where as 3.5% of cacnelations were returning guests; 

42672 / (43672 + 552)



#then, we can look at customer segmentation(which country they come from, how they booked; once we do that, we can once again relate to this cancelations); perhaps we can see what type of families are coming to these hotels as well 

top_countries <- df %>% group_by(country) %>% count(country) %>% arrange(desc(n))
top <- top_countries[1:20,]
ggplot(top, aes(x = country , y = n)) + geom_bar( stat = 'identity')  ##We can see most bookings come from portual or other european countries; makes sense considering these hotels are booked in europe 



###market segmentation 
ggplot(df, aes(x = factor(1), fill = market_segment)) + geom_bar(width = 1) + coord_polar(theta = "y") + theme_void()

ggplot(df, aes(x = market_segment, fill = factor(is_canceled))) + geom_bar() ###it seems that groups market_segment is the most likely to cancel, while direct is the least likeliy to cancel 


##customer types 
ggplot(df, aes(x = factor(1), fill = customer_type)) + geom_bar(width = 1) + coord_polar(theta = "y") + theme_void()
ggplot(df, aes(x = customer_type, fill = is_canceled)) + geom_bar() ###transient customers most likley to cancel 


##we can also find the probabilities of each group being likely to cancel 
df %>% group_by(market_segment) %>% summarise(ProbabilityofCancelation = mean(is_canceled)) %>% ggplot(aes(x= market_segment , y = ProbabilityofCancelation)) + geom_bar(stat = 'identity')

##there is variation, so we can definitely include this in our model 

######################################

#PRICE RELATIONSHIPS 

##once we do this, we can look at relationship sbetween type of room and price; then we can see if there's a relationship between price and cancelation or tyoe of room and cancelation 


##let's also observe the distribution of stay in weekend nights vs stay in week nights; how do adr 
df

##I would like to know if prices tend to be higher on stays_in_wekedn_nights vs week nights; how exactly will this help our model? 
##It proabbly won;t 

##essentially, the goal is to see how adr will affect changes in booking prices; if adr will affect booking prices, then this has taken into account stays in weekend nights

##suppose we have a high number of stays; then adr will likely be smaller? and if it's smaller, then 


##adr versus number of stays; 
plot(data = df, adr ~ stays_in_weekend_nights + stays_in_week_nights)  

##we can see that these follow nearly identical strends; the average daily rate for each given stay, whether in week nights or weekend nights, does not seem to affect 
##adr that much; 


##but let's see if how much adr differs based on bookings vs cancellations; the question is: does knowing there is a higher pricing for one hotel booking mean
#that someone is less likely to cancel? 


ggplot(data =df, aes(x = factor(is_canceled),  y = adr)) + geom_boxplot()
##there's an outlier in this plot that distorts our boxplot; we will remove the outlier to get a better picture 

df_no_outlier <- df[-which(df$adr == max(df$adr)),]

ggplot(data =df_no_outlier, aes(x = factor(is_canceled),  y = adr)) + geom_boxplot(col = "blue")


##on average, it seems like people who canceled tend to have a higher adr; however, let's conduct a t.test to see if there differences in means is significant 
t.test(df$adr ~ df$is_canceled, var.equal = TRUE)

##according to this, there is a statistically significant difference given the means. So, even though our visualization does show a strong relationship, we will use adr in our model. 


##we also want to see some distinctions between the two hotels; the purpose of this task is to identify the general trends and factors leading to drops in bookign times; 
##of course, we want to see the differences in the two hotels;  we will include this in our model 


##Our goal is to generalize and create a model for all hotel bookings, but to do this, we have to take into account differences between the two hotels give to us. 


ggplot(data = df, aes(x = hotel, fill = factor(is_canceled))) + geom_bar()

## we have many more data points from the city hotel; relatively, we can see that more people tend to cancel with the city hotel; 

##let's look at price differences between the city hotel and resort hotel 

ggplot(data = df_no_outlier, aes(x = assigned_room_type, y = adr, fill = hotel)) + geom_boxplot()

##given the room type, on average we see city hotels are higher for most room types; 
df %>% group_by(assigned_room_type) %>% summarise(avgprice = mean(adr)) %>% ggplot(aes(x = assigned_room_type, y = avgprice)) + geom_bar(stat = 'identity', fill = 'orange')



```



Classification 

Lgistic regression 
```{r}


df$is_canceled = as.factor(df$is_canceled)

##training/testing model 
smp_size <- floor(0.75 * nrow(df))

## set the seed to make your partition reproducible
set.seed(123)
train_ind <- sample(seq_len(nrow(df)), size = smp_size)

train <- df[train_ind, ]
test <- df[-train_ind, ]

model <- glm(data = train, is_canceled ~ adr +  lead_time+ arrival_date_month + adults + children + babies + hotel, family = "binomial")

summary(model)


##Assessing our model
##we can see the most important predictors here 
anova(model, test = "Chisq") 

##null model is only the model with the intercept; 
##higher rresdiaul deviance means your model is doing better; 

probs <- predict(model, test[-2], type = "response" )

our.pred = rep("0", dim(test)[1])
our.pred[our.pred > .5] = "1"

mean(our.pred == train$is_canceled)
misClasificError <- mean(fitted.results != test$Survived)

vif(model)
plot(test$is_canceled, our.pred)
summary(model)



```


random forest 

```{r}

##some packages for this model 
library(cowplot) 
library(randomForest)


df<-df %>% select(-c("arrival_date_year","company" , "arrival_date_week_number", "arrival_date_day_of_month", "agent","days_in_waiting_list", "reservation_status", "reservation_status_date"))

df <- df %>% select(-c("country"))
df <- df %>% select(-c("meal", "distribution_channel"))
df$is_canceled <- as.factor(df$is_canceled)

levels(df$is_canceled) <- c("Not Canceled", "Canceled")
df[which(is.na(df)),]  ## only 4 NAs, so we will omit the missing valules 
df<- na.omit(df)##Data without NAs


model <- randomForest(is_canceled ~. ,data = df[1:2000,], ntree = 1000,  proximity = TRUE) 
?randomForest
#Returns the proximity matrix
model
#notice classifciation is default, since we are predicting categorical variable
#number of trees is 500, no of variabels at each split is 3 (by default)


#OOB error estimate

#83.5% of ourt of bag samples were correctly classified by random forest 

#confusion matrix to show results  
randomForest

oob.error.data <- data.frame(
  Number_of_Trees=rep(1:nrow(model$err.rate), times=3),
  Type=rep(c("OOB", "Not Canceled", "Canceled"), each=nrow(model$err.rate)),
  Error_Type=c(model$err.rate[,"OOB"], 
    model$err.rate[,"Not Canceled"], 
    model$err.rate[,"Canceled"]))

ggplot(data=oob.error.data, aes(x=Number_of_Trees, y=Error_Type)) +
  geom_line(aes(color=Type)) + ggtitle("Plotting Errors for each Tree")
##after 600 trees, variability seems to drop quite a bit;thus, optimal number of trees we'll choose 600 


##now we will use cross validation to find the optimal number of trees 
##mtry is the number of selections per each cut in the trees 
oob.values <- vector(length=10)
for(i in 1:10) {
  temp.model <- randomForest(is_canceled ~. ,mtry = i, data = df[1:2000,], ntree = 600) 
  oob.values[i] <- temp.model$err.rate[nrow(temp.model$err.rate),1]
}

oob.values[which(oob.values == min(oob.values))]
##20.2 is the lowest out of bag error we get 

```

